---
title: "Visualizing Maps"
toc: false
format:
  html:
    code-fold: true
---

#### A project that uses mapping features from different libraries in R to create static and interactive choropleth plots.

```{r,message=FALSE,include=FALSE}
# data
library(tidyverse)
setwd("~")
PLACES <- read_csv("R/SDS264_F24/HW/Projects/Data/PLACES.csv")
PLACES <- PLACES |>
  rename(State = StateDesc)

# getting state shapes
library(maps)
states <- map_data("state")
head(states)
```


## First Map: Prevalence of Depression Across the United States


```{r, fig.alt="This is a choropleth map or plot of the United States. Each state is drawn out and colored according to its average prevalence of depression, which ranges from 16 to 29 percent. The states with the lowest rates of depression appear to be Nebraska and New Jersey, while the highest prevalence of depression occurs in Tennessee and West Virginia. In general, the pattern of prevalence of depression does not seem to be in a relationship with political trends, rural versus urban dynamics, nor coastal versus mainland geographical locations.", fig.width=8}
# First

# summary by state
PLACES_states <- PLACES |>
  mutate(State = tolower(State)) |>
  group_by(State) |>
  summarize(
    continuous = mean(DEPRESSION_CrudePrev), 
    other = mean(SLEEP_CrudePrev)
  )

# join
chimera <- states |>
  left_join(PLACES_states, by = c("region" = "State"))

# plot
chimera |>
  ggplot(aes(x = long, y = lat, group = group)) +
    geom_polygon(aes(fill = continuous), color = "#222200") +
    coord_map() +  
    theme_void() +  
    scale_fill_viridis_c(option = "E") +
    theme(panel.background = element_rect(fill = "beige")) +
    labs(
      title = "Average Prevalence of Depression by State",
      fill = "Mean Prevalence of Depression",
      caption = "Source: CDC's PLACES Census Tract Data 2024 Release\nhttps://data.cdc.gov/500-Cities-Places/PLACES-Census-Tract-Data-GIS-Friendly-Format-2024-/yjkw-uj5s"
    )
```


## Second Map: Prevalence of Lack of Sleep Across the United States


```{r, fig.width=8}
# First
AVG <- mean(PLACES_states$other)

chimera <- chimera |>
  mutate(categorical = ifelse(other>AVG, "More Than Average", "Less Than Average"))

# plot
chimera |>
  ggplot(aes(x = long, y = lat, group = group)) +
    geom_polygon(aes(fill = fct(categorical)), color = "#353535") +
    coord_map() +  
    theme_void() +  
    scale_fill_viridis_d(option = "F") +
    theme(panel.background = element_rect(fill = "lavender")) +
    labs(
      title = "U.S. States With Higher Than Average Sleep Deprivation",
      subtitle = "(where the average prevalence of sleep deprivation is 35.82%)",
      fill = "Prevalence of Sleep Deprivation",
      caption = "Source: CDC's PLACES Census Tract Data 2024 Release\nhttps://data.cdc.gov/500-Cities-Places/PLACES-Census-Tract-Data-GIS-Friendly-Format-2024-/yjkw-uj5s"
    )
```


**Description:**
This plot highlights the geographical divide between the states with a higher prevalence of sleep deprivation than average and those with a lower prevalence. The prevalence for each state is calculated by averaging the percentage across its counties; by splitting the states into two categories, it becomes extremely clear that the Southeast as a geographical region has a higher prevalence of lack of sleep. This does not seem to follow the trends observed in the map of depression; however, it would be premature to assume that sleep deprivation and depression are not correlated, as there could be many other factors contributing to the trends observed in these maps.




```{r,message=FALSE, include=FALSE}
# data

library(tidyverse)
setwd("~")
PLACES <- read_csv("R/SDS264_F24/HW/Projects/Data/PLACES.csv")
AVG <- mean(PLACES$SLEEP_CrudePrev)

PLACES <- PLACES |>
  rename(name = StateDesc) |>
  group_by(name) |>
  summarize(
    depr = mean(DEPRESSION_CrudePrev), 
    sleep = mean(SLEEP_CrudePrev)
  ) |>
  ungroup() |>
  mutate(sleep = ifelse(sleep>AVG, "Above average", "Below average")) |>
  mutate(sleep = fct(sleep))
  

library(sf) 
states <- read_sf("https://rstudio.github.io/leaflet/json/us-states.geojson")


# joined
PLACES_states <- states |>
  left_join(PLACES, by = c("name" = "name"))

library(leaflet)
```


## First Map: Interactive

```{r, message = FALSE}
# First

# bins
bins <- c(15, 17.5, 20, 22.5, 25, 27.5, 30)
pal <- colorBin("plasma", domain = PLACES_states$depr, bins = bins)

# labels
library(htmltools)
library(glue)

PLACES_states <- PLACES_states |>
  mutate(labels = str_c("<u>", name, "</u>: ", round(depr,2), "% prevalence"))

labels <- lapply(PLACES_states$labels, HTML)


# plot
leaflet(PLACES_states) |>
  setView(-96, 37.8, 4) |>
  addTiles(attribution = "Source: <a href='https://data.cdc.gov/500-Cities-Places/PLACES-Census-Tract-Data-GIS-Friendly-Format-2024-/yjkw-uj5s'>CDC PLACES</a>") |>
  addPolygons(
    fillColor = ~pal(depr),
    weight = 1,
    opacity = 1,
    color = "black",
    dashArray = "",
    fillOpacity = .9,
    highlightOptions = highlightOptions(
      weight = 5,
      color = "white",
      dashArray = "",
      fillOpacity = .5,
      bringToFront = TRUE),
    label = labels,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "15px",
      direction = "auto")) |>
  addLegend(pal = pal, values = ~density, opacity = 0.7, title = "Prevalence of Depression",
    position = "bottomright")
```

# Second Map: Interactive

```{r, message = FALSE}
# Second

# bins
pal <- colorFactor("magma", domain = PLACES_states$sleep, levels = c("Above average", "Below average"))

# labels
library(htmltools)
library(glue)

PLACES_states <- PLACES_states |>
  mutate(labels = str_c("<b>", name, "</b>: ", sleep, " prevalence of sleep deprivation"))

labels <- lapply(PLACES_states$labels, HTML)


# plot
leaflet(PLACES_states) |>
  setView(-96, 37.8, 4) |>
  addTiles(attribution = "Source: <a href='https://data.cdc.gov/500-Cities-Places/PLACES-Census-Tract-Data-GIS-Friendly-Format-2024-/yjkw-uj5s'>CDC PLACES</a>") |>
  addPolygons(
    fillColor = ~pal(sleep),
    weight = 1,
    opacity = 1,
    color = "darkgray",
    dashArray = "",
    fillOpacity = .6,
    highlightOptions = highlightOptions(
      weight = 3,
      color = "darkred",
      dashArray = "",
      fillOpacity = 1,
      bringToFront = TRUE),
    label = labels,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "15px",
      direction = "auto")) |>
  addLegend(pal = pal, values = PLACES_states$sleep, opacity = 0.7, 
          title = "Prevalence of Sleep Deprivation", position = "bottomright")
```







